---
title = "Cache"
---

Documents with expensive code chunks (model fitting, simulations, large data imports) can take a long time to render. Since most edits only touch prose or a single chunk, re-executing every chunk from scratch is wasteful and slow. This is especially painful with `calepin --preview`, which watches the source file and re-renders on every save, ideally instantly, without waiting for expensive code chunks to run again.

*Calepin* solves this by caching code chunk results. Unchanged chunks are skipped entirely on subsequent renders, reducing rebuild times from minutes to milliseconds. Caching is on by default.

Disable caching for a single chunk with `#| cache = false`:

````markdown
```{r}
#| cache = false
slow_function()
```
````

Disable caching for the entire document in TOML front matter:

```toml
---
[execute]
cache = false
---
```

Cached results are stored in `.calepin/cache/{stem}/` (see [Storage layout](#storage-layout) for details). Changing a chunk's source code, or any chunk above it, automatically invalidates the cache (see [Upstream digest chain](#upstream-digest-chain)).

# Example

The two chunks below each sleep for 2 seconds to simulate expensive computation. On first render, total execution takes about 4 seconds. On second render, both hit the cache and finish in under a second.

```{r, slow-r}
Sys.sleep(2)
x <- sum(1:1000)
x
```

```{python, slow-py}
import time
time.sleep(2)
y = sum(range(1, 1001))
print(y)
```

Render this document twice and compare timings:

```bash
time calepin cache.qmd     # first run: ~4s (both chunks execute)
time calepin cache.qmd     # second run: <1s (both chunks cached)
```

# Design

The caching design is inspired by the approach in the excellent [litedown package for `R`](https://yihui.org/litedown/) by [Yihui Xie](https://yihui.org/), who pioneered chunk-level caching with upstream invalidation in R Markdown tooling. *Calepin* adapts that design to a Rust pipeline operating across both R and Python engines.

## Cache level

Caching operates at the `ChunkResult` level, after code execution but before results are converted into display elements. A `ChunkResult` is the raw engine output: printed text, warnings, messages, errors, and plot file paths. This is the natural seam because it captures everything the subprocess produced while remaining independent of display options. Changing `echo`, `include`, `results`, `warning`, `message`, or `comment` reuses the cache; changing source code, `fig.width`, or `dev` invalidates it.

## Cache key

Each chunk's cache key is an xxh3-128 hash of:

| Component | Rationale |
|-----------|-----------|
| Source code | Code changed means re-execute |
| All non-display chunk options | Any option that affects execution output invalidates the cache |
| Upstream digest | Cumulative hash of all prior chunk keys in document order |

Display-only options (`echo`, `include`, `results`, `warning`, `message`, `comment`, `fig-cap`, `fig-alt`, `fig-cap-location`, `tbl-cap`, `cache`) are excluded because they control how results are shown, not what the engine produces. All other options are included by default, so new options invalidate the cache without needing to update the caching module.

## Upstream digest chain

R and Python sessions are stateful: each chunk can define variables, load libraries, or modify globals that later chunks depend on. If chunk 3 changes, chunks 4 and onward must re-execute even if their own source is unchanged.

The upstream digest solves this. It is a rolling xxh3-128 hash that incorporates every chunk's key in document order. Before computing its own key, each chunk mixes the current upstream digest into the hash. After computing the key, the chunk advances the digest:

```
for each chunk:
    key = xxh3(source, options, upstream_digest)
    upstream_digest = xxh3(upstream_digest, key)
```

This means any upstream change, even in a non-cached chunk, cascades invalidation to all downstream cached chunks. Inline code expressions also advance the digest, since they can mutate session state.

Chunks with `eval: false` still contribute to the digest chain (based on their source and options) so that toggling `eval` correctly invalidates downstream caches.

## Storage layout

Cache files live in `.calepin/cache/{stem}/{label}/`:

```
.calepin/cache/tutorial/
  slow-r/
    meta.json          # {"hash": "<hex>", "plot_files": ["slow-r-1.png"]}
    results.bincode    # serialized Vec<ChunkResult>
    slow-r-1.png       # cached plot copy
  slow-py/
    meta.json
    results.bincode
```

On a cache hit, plot files are copied back from the cache directory to the figure output directory so that downstream rendering finds them at the expected paths. On a miss, the chunk executes normally and results plus plots are written into the cache directory.

## Session state on cache hit

When a cached chunk is skipped, its code never runs in the R/Python subprocess, so variables it defines are not available to later chunks. For example:

````markdown
```{r, fit-model}
model <- lm(mpg ~ wt, data = mtcars)
summary(model)
```

```{r, use-model}
#| cache = false
predict(model, newdata = data.frame(wt = 3))
```
````

On the first render, both chunks execute and everything works. On the second render, `fit-model` is cached and skipped, so `model` is never defined in the R session. The `use-model` chunk (which opted out of caching) will fail because `model` does not exist.

This is typically not a problem when caching is left at its default (all chunks cached), since cached chunks only see cached output, not live session state. The issue arises when individual chunks opt out with `#| cache = false` and depend on variables from cached chunks above them.

