---
title: Chunks and Inline Code
---

*Calepin* embeds R, Python, and shell runtimes directly in the binary, so code chunks are executed during rendering without spawning external processes. Variables persist across chunks within the same language, plots are captured automatically, and results are woven into the final document.

# Code chunks

A code chunk has four parts:

````
```{r, my-chunk}
#| echo: false
#| fig-width: 7
x <- 1:10
mean(x)
```
````

1. **Language** --- The header starts with the language engine: `{r}`, `{python}`, or `{sh}`.
2. **Label** (optional) --- After the language, separated by a comma: `my-chunk`. Labels identify chunks in error messages and are used for cross-referencing figures (see [Figures](../figures.qmd)).
3. **Chunk options** --- Lines starting with `#|` inside the chunk set options in `key: value` format. Option names use dashes, not dots (e.g., `fig-width` not `fig.width`). See [Chunk options] below for the full list.
4. **Code** --- The body of the chunk is executed by the language engine. Output is captured and inserted into the document.


# Inline code

Inline expressions are evaluated by wrapping them in backticks with the engine name. The mean of 1:10 is `{r} mean(1:10)` and pi is approximately `{r} round(pi, 4)`.

Numeric results in R are automatically formatted with 3 significant digits and comma separators for large numbers.

Python and shell inline expressions work the same way: `{python} 2 + 2` and `{sh} echo hello`.


# Multiple expressions

When a chunk contains several expressions, each produces its own output block, interleaved with the source.

```{r}
cat("Line 1\n")
cat("Line 2\n")
1 + 1
```


# Chunk options

Options are set with `#|` pipe comments inside the chunk. Each option goes on its own line in `key: value` format. Option names use dashes, not dots (e.g., `fig-width` not `fig.width`).

## echo

Controls whether the source code is displayed. Default: `true`.

```{r}
#| echo: false
summary(mtcars$mpg)
```

## eval

Controls whether the code is executed. Default: `true`. Set to `false` to display code without running it.

```{r}
#| eval: false
stop("This error is never reached")
```

## include

Controls whether the chunk produces any output at all. When `false`, the code runs silently (useful for setup chunks). Default: `true`.

```{r, setup}
#| include: false
library(stats)
```

## results

Controls how text output is handled. Options: `markup` (default), `asis` (inject raw output into the document), `hide` (suppress output).

```{r}
#| results: hide
invisible_result <- 42
```

The value was computed silently: `{r} invisible_result`.

## warning and message

Control whether warnings and messages from R are displayed. Default: `true` for both.

```{r}
#| warning: false
#| message: false
log(-1)
```


# Option reference

| Option | Default | Description |
|--------|---------|-------------|
| `eval` | `true` | Execute the code |
| `echo` | `true` | Display the source code |
| `include` | `true` | Include any output (if `false`, runs silently) |
| `results` | `markup` | Output handling: `markup`, `asis`, or `hide` |
| `warning` | `true` | Display warnings |
| `message` | `true` | Display messages |
| `fig-width` | 7 | Figure width in inches (see [Figures](../figures.qmd)) |
| `fig-height` | 5 | Figure height in inches (see [Figures](../figures.qmd)) |
| `fig-cap` | | Figure caption (see [Figures](../figures.qmd)) |
| `fig-alt` | | Figure alt text (see [Figures](../figures.qmd)) |
| `dev` | `png` | Graphics device (see [Figures](../figures.qmd)) |


# Syntax highlighting

*Calepin* uses [syntect](https://github.com/trishume/syntect) for syntax highlighting. Set the theme with `highlight-style` in the front matter (Quarto-compatible key name):

```yaml
---
highlight-style: solarized-dark
---
```

The default is `inspired-github`. To list all built-in themes:

```
calepin --list-highlight-styles
```

You can also point to a custom `.tmTheme` file:

```yaml
---
highlight-style: mytheme.tmTheme
---
```

For HTML output, highlighting produces CSS classes, so the theme controls only the generated stylesheet. For LaTeX, colors are inlined via `\textcolor`. Typst and Markdown output receive no highlighting.
