---
title = "Chunks"
---

*Calepin* runs R, Python, and shell code via persistent subprocesses (`Rscript`, `python3`, `/bin/sh`). Variables persist across chunks within the same language, plots are captured automatically, and results are woven into the final document.

{% include "assets/source_tip.html" %}

# Code chunks

A code chunk has four parts:

1. Language --- The header starts with the language engine: `{r}`, `{python}`, or `{sh}`.
2. Label (optional) --- After the language, separated by a comma: `my-chunk`. Labels identify chunks in error messages and are used for cross-referencing figures (see [Figures](../authoring/figures.qmd)).
3. Chunk options --- Lines starting with `#|` inside the chunk set options in `key = value` format (TOML). For Quarto compatibility, `key: value` also works. Option names use dashes, not dots (e.g., `fig-width` not `fig.width`). See [Chunk options] below for the full list.
4. Code --- The body of the chunk is executed by the language engine. Output is captured and inserted into the document.

Here is a simple example. The `echo: fenced` option shows the chunk exactly as written in the source, including the fence markers and options:

```{r, my-chunk}
#| echo = fenced
x <- 1:10
mean(x)
```


# Multiple expressions

When a chunk contains several expressions, each produces its own output block, interleaved with the source:

```{r}
#| echo = fenced
cat("Line 1\n")
cat("Line 2\n")
1 + 1
```


# Chunk options

Options are set with `#|` pipe comments inside the chunk. Each option goes on its own line in `key = value` format (TOML). For Quarto compatibility, `key: value` also works. Option names use dashes, not dots (e.g., `fig-width` not `fig.width`).

## echo

Controls whether the source code is displayed. Default: `true`. When set to `fenced`, the chunk is shown with its fence markers and options --- useful for tutorials.

```{r}
#| echo = fenced
#| eval = false
summary(mtcars$mpg)
```

Setting `echo = false` hides the chunk entirely. Only the output is shown:

```{r}
#| echo = false
summary(mtcars$mpg)
```

## eval

Controls whether the code is executed. Default: `true`. Set to `false` to display code without running it:

```{r}
#| echo = fenced
#| eval = false
stop("This error is never reached")
```

## include

Controls whether the chunk produces any output at all. When `false`, the code runs silently (useful for setup chunks). Default: `true`.

```{r, setup}
#| echo = fenced
#| include = false
library(stats)
```

## results

Controls how text output is handled. Options: `markup` (default), `asis` (inject raw output into the document), `hide` (suppress output).

```{r}
#| echo = fenced
#| results = hide
invisible_result <- 42
```

The value was computed silently: `{r} invisible_result`.

## warning and message

Control whether warnings and messages from R are displayed. Default: `true` for both.

```{r}
#| echo = fenced
#| warning = false
#| message = false
log(-1)
```


# Option reference

| Option | Default | Description |
|--------|---------|-------------|
| `eval` | `true` | Execute the code |
| `echo` | `true` | Display the source code (`true`, `false`, or `fenced`) |
| `include` | `true` | Include any output (if `false`, runs silently) |
| `results` | `markup` | Output handling: `markup`, `asis`, or `hide` |
| `warning` | `true` | Display warnings |
| `message` | `true` | Display messages |
| `fig-width` | 6 | Figure width in inches (see [Figures](../authoring/figures.qmd)) |
| `fig-height` | derived | Derived from `fig-width * fig-asp` (see [Figures](../authoring/figures.qmd)) |
| `fig-cap` | | Figure caption (see [Figures](../authoring/figures.qmd)) |
| `fig-alt` | | Figure alt text (see [Figures](../authoring/figures.qmd)) |
| `dev` | `png` | Graphics device (see [Figures](../authoring/figures.qmd)) |


# Syntax highlighting

*Calepin* uses [syntect](https://github.com/trishume/syntect) for syntax highlighting. Themes are controlled by color schemes, not a standalone setting. Each color scheme bundles a light and dark syntax theme pair.

Set the color scheme in `config.toml` or front matter:

```toml
colors = ["nord", "ayu", "catppuccin-mocha"]
```

The first entry is the default. All entries populate the color picker widget on websites. Built-in schemes: `nord` (default), `ayu`, `black`, `catppuccin-frappe`, `catppuccin-macchiato`, `catppuccin-mocha`.

To list all built-in color schemes:

```
calepin extra colors
```

For HTML output, highlighting produces CSS classes, so the theme controls only the generated stylesheet. For LaTeX, colors are inlined via `\textcolor`. Typst and Markdown output receive no highlighting.
