---
title = "Figures"
bibliography = "assets/library.bib"
---

Code chunks that produce plots automatically generate figures. Use `fig-cap` for a caption and a label in the chunk header for cross-referencing.

```{r, fig-scatter}
#| echo = fenced
#| fig-cap = "A simple scatter plot"
plot(
    1:10,
    (1:10)^2,
    main = "Square Numbers",
    xlab = "x",
    ylab = "x^2"
)
```

Figures with labels can be cross-referenced with `@fig-label` (see @fig-scatter above).

Any plotting library works. Here is a ggplot2 example:

```{r, fig-ggplot}
#| echo = fenced
#| fig-cap = "Fuel efficiency by weight (ggplot2)"
library(ggplot2)
ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  labs(x = "Weight (1000 lbs)", y = "MPG", color = "Cylinders") +
  theme_minimal()
```


# Size

In Quarto, figure appearance is governed by four parameters: `fig-width`, `fig-height`, `fig-asp`, and `out-width`. Unless users pick exactly the right combination, figures end up with inconsistent font sizes, axis labels, and line widths. A narrow figure has tiny text while a wide figure has huge text, because the graphics device size changes but the display size does not (or vice versa).

*Calepin* uses pragmatic defaults and adjusts some parameters automatically to maintain a consistent look. In practice, we normally only need two options:

| Option | Default | What it controls |
|--------|---------|------------------|
| `out-width` | `70%` | Horizontal space in the document |
| `fig-asp` | `0.618` | Aspect ratio (golden ratio) |

*Calepin* automatically scales the internal graphics device (`fig-width`) to match `out-width`, so text and labels inside figures stay the same size regardless of how wide the figure is displayed. This means a narrow figure and a wide figure will have consistent font sizes.

* Want a wider figure? Change `out-width`. 
* Want a taller figure? Change `fig-asp`. 

That is all.

```{r}
#| fig-asp = 0.4
#| out-width = "95%"
#| fig-cap = "A wide, short figure"
#| echo = fenced
plot(1:20, sin(1:20), type = "l", col = "steelblue", lwd = 2)
```

The `fig-width` and `fig-height` options still exist for full manual control, but you rarely need them. When you set `fig-width` explicitly, auto-scaling is disabled.


# Options

| Option | Default | Description |
|--------|---------|-------------|
| `out-width` | `70%` | Display width in the document |
| `fig-asp` | `0.618` | Aspect ratio (height / width) |
| `fig-width` | `6` | Graphics device width in inches (auto-scaled from `out-width`) |
| `fig-height` | | Derived from `fig-width * fig-asp` |
| `fig-cap` | | Caption text |
| `fig-alt` | | Alt text for accessibility |
| `fig-align` | `center` | Alignment (`left`, `center`, `right`) |
| `dev` | `png` | Graphics device (`png`, `pdf`, `svg`) |


# Layouts

Figures can be arranged in grids using `layout-ncol`. The following example places three plots side by side with individual subcaptions and an overall caption.

{% include "source_tip" %}

::: {#fig-panels layout-ncol=3 fig-cap="Three growth patterns compared."}
```{r, line}
#| echo = false
#| out-width = "100%"
#| fig-width = 3
#| fig-height = 3
#| fig-cap = "Linear"
plot(1:10, 1:10, type = "l", col = "steelblue", lwd = 2)
```

```{r, quad}
#| echo = false
#| out-width = "100%"
#| fig-width = 3
#| fig-height = 3
#| fig-cap = "Quadratic"
plot(1:10, (1:10)^2, type = "l", col = "firebrick", lwd = 2)
```

```{r, log}
#| echo = false
#| out-width = "100%"
#| fig-width = 3
#| fig-height = 3
#| fig-cap = "Logarithmic"
plot(1:10, log(1:10), type = "l", col = "forestgreen", lwd = 2)
```
:::

See @fig-panels for the result. The `layout` attribute supports custom grids with relative widths (e.g., `layout="[[1,1],[1]]"`) and negative values for spacing gaps.
