---
title: Figures
bibliography: ../library.bib
---

::: {.callout-tip}
Click the `</>` button in the top navigation bar to view the original `.qmd` source code.
:::

Code chunks that produce plots automatically generate figures. Use `fig-cap` for a caption and a label in the chunk header for cross-referencing.

```{r, scatter}
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "A simple scatter plot"
plot(
    1:10,
    (1:10)^2,
    main = "Square Numbers",
    xlab = "x",
    ylab = "x^2"
)
```

Figures with labels can be cross-referenced with `@fig-label`. See @fig-scatter above.

Any plotting library works — here is a ggplot2 example:

```{r, fig-ggplot}
#| fig-width: 6
#| fig-height: 4
#| fig-cap: "Fuel efficiency by weight (ggplot2)"
library(ggplot2)
ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  labs(x = "Weight (1000 lbs)", y = "MPG", color = "Cylinders") +
  theme_minimal()
```


# Figure options

| Option | Default | Description |
|--------|---------|-------------|
| `fig-width` | 7 | Width in inches |
| `fig-height` | 5 | Height in inches |
| `fig-cap` | | Caption text |
| `fig-alt` | | Alt text for accessibility |
| `dev` | `png` | Graphics device (`png`, `pdf`, `svg`) |


# Figure layouts

Figures can be arranged in grids using `layout-ncol`. The following example places three plots side by side with individual subcaptions and an overall caption.

::: {#fig-panels layout-ncol=3}

```{r, line}
#| echo: false
#| fig-cap: "Linear"
plot(1:10, 1:10, type = "l", col = "steelblue", lwd = 2)
```

```{r, quad}
#| echo: false
#| fig-cap: "Quadratic"
plot(1:10, (1:10)^2, type = "l", col = "firebrick", lwd = 2)
```

```{r, log}
#| echo: false
#| fig-cap: "Logarithmic"
plot(1:10, log(1:10), type = "l", col = "forestgreen", lwd = 2)
```

Three growth patterns compared.
:::

See @fig-panels for the result. The `layout` attribute supports custom grids with relative widths (e.g., `layout="[[1,1],[1]]"`) and negative values for spacing gaps.


# Figure templates

The figure element template controls how each figure is rendered. The built-in template uses these variables:

| Variable | Description |
|----------|-------------|
| `{{path}}` | Image file path |
| `{{alt}}` | Alt text |
| `{{caption}}` | Figure caption |
| `{{label}}` / `{{number}}` | Label and number |
| `{{image}}` | Pre-rendered image markup |

To customize figure output, create `_calepin/elements/figure.html` (or `.tex`, `.typ`, `.md`) with these placeholders.


# Lightbox

The `lightbox` plugin wraps images inside `.figure` divs with click-to-enlarge overlays. Clicking an image opens it full-screen; clicking the overlay dismisses it. Only applies to HTML output.

```yaml
calepin:
  plugins:
    - lightbox
```

No per-image configuration needed---all figure images get lightbox behavior automatically.
