---
title: Tables
---

# Markdown

Tables can use the standard Github-Flavored Markdown pipe syntax. Column alignment is controlled by colons in the separator row: left (default), right (`---:`), or center (`:---:`).

| Name  | Score | Grade |
|-------|------:|:-----:|
| Alice |    95 | A     |
| Bob   |    87 | B     |
| Carol |    92 | A     |


## Captions and cross-references

Wrap a table in a fenced div with a `#tbl-` id to add a caption and make it cross-referenceable. The last paragraph in the div becomes the caption:

```
::: {#tbl-grades}
| Name  | Score | Grade |
|-------|------:|:-----:|
| Alice |    95 | A     |
| Bob   |    87 | B     |
| Carol |    92 | A     |

Student grades for Fall 2026.
:::
```

::: {#tbl-grades}
| Name  | Score | Grade |
|-------|------:|:-----:|
| Alice |    95 | A     |
| Bob   |    87 | B     |
| Carol |    92 | A     |

Student grades for Fall 2026.
:::

Reference tables with `@tbl-label`: @tbl-grades shows the full class results. Suppress the type prefix with `[-@tbl-label]`: [-@tbl-grades].


## Caption location

By default, captions appear above the table. Use `tbl-cap-location` to place them below:

::: {#tbl-bottom tbl-cap-location="bottom"}
| x | y  |
|---|---:|
| 1 |  1 |
| 2 |  4 |
| 3 |  9 |

This caption appears below the table.
:::


# R

The [tinytable](https://vincentarelbundock.github.io/tinytable/) R package generates format-aware tables that work automatically in *Calepin*. Use a `tbl-` label on the chunk and `tbl-cap` for the caption. The `style_tt()` function applies cell-level formatting, and `group_tt()` adds spanning headers.

```{r, tbl-cars}
#| tbl-cap: Motor Trend car data
#| echo: fenced
library(tinytable)
tt(mtcars[1:5, 1:4]) |>
  style_tt(2:3, background = "#5E81AC", color = "white") |>
  group_tt(j = list("Group" = 3:4))
```

Cross-reference work as expected, by referring to the figure label: see @tbl-cars.

The [gt](https://gt.rstudio.com/) and [`knitr`](https://yihui.org/knitr/) are two other popular packages in `R` to generate tables:

```{r, tbl-cars}
#| tbl-cap: Motor Trend car data
#| echo: fenced
gt::gt(mtcars[1:5, 1:4])
```

```{r, tbl-cars}
#| tbl-cap: Motor Trend car data
#| echo: fenced
knitr::kable(mtcars[1:5, 1:4])
```
