---
title: "Nesting with data.table and tidyverse"
author: "Vincent Arel-Bundock"
date: "2020-08-17"
---

The `nest`/`unnest` programming pattern can be a useful way to organize your data analyses in `R`. The idea is to store a series of objects (e.g., datasets, models, summaries) in a "list-column." A list-column is simply a column in your data frame where each entry is a list. Each of the lists in your list-column can itself contain any object you want, including other datasets or statistical model objects. The benefit of a list columns is that analysts can operate on them using standard tools such as `lapply` or `purrr::map`. 

One common use-case is to split a dataset into subgroups, and to "nest" those subgroups of the data into a list-column. Then, we can estimate models and summarize the data easily by operating on that list column.

I could not find an easy introductory tutorial for nesting list-columns in `data.table`, so I decided to write this notebook. In it, I compare how list-columns can be created in both `data.table` and `tidyverse`. 

To begin, we load libraries and download the Guerry data from [Rdatasets](vincentarelbundock.github.io/Rdatasets/):

```{r}
#| message: false
library(data.table)
library(tidyverse)
library(broom)
library(modelsummary)
library(microbenchmark)

url <- "https://vincentarelbundock.github.io/Rdatasets/csv/HistData/Guerry.csv"
dat <- fread(url)
dat <- na.omit(dat)
```

## `data.table`

In `data.table`, we can split our dataset using the `by` argument, and return any summary of the data we want inside a list object. For instance, we calculate the standard deviation of the "Clergy" variable for each region by typing:

```{r}
dat[, list(clergy_sd=sd(Clergy)), by=Region]
```

To create a list-column, we use the same syntax, but instead of returning a single numeric value (`sd(Clergy)`), we return a list with the ["Subset of Data"](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-sd-usage.html) special object (`.SD`) inside:

```{r}
dat = dat[, list(data=list(.SD)), by=Region]
```

The result is a data table with a new list-column called `data`.  Each element of the `data` column is now a list, with a data table inside, and each of those data tables has 17 rows and 23 columns:

```{r}
dat
```

This is a convenient data structure, because it allows us to operate on the data subsets with standard commands like `lapply` or `purrr::map`. For instance, this estimates a separate linear regression model in each subset of the data:

```{r}
dat[, model := lapply(data, function(x) lm(Donations ~ Clergy, x))]
dat
```

Above, we see that the dataset now includes a new list-column called "model". Each element of that column is a list, with a `lm` object inside.  We can do cool things with this list-column. For example, we can pass it directly to `modelsummary`: 

```{r}
modelsummary(dat$model)
```

We could also use the `broom::tidy` function to extract model estimates in a tidy format:

```{r}
dat[, result := lapply(model, tidy)]
dat
```

In this new "result" list-column, each entry is a list with a data frame-like object inside (a `tibble`). This data frame includes information about coefficient estimates, standard errors, etc. To extract that information from the list-column we move to the final step of today's demonstration: unnesting.

Since each element of the `result` column is a list with a single data frame inside, we simply extract that single element with brackets `[[1]]`:

```{r}
dat[, result[[1]], by=Region]
```

Putting everything together, we have:

```{r}
#| eval: false
# Nest
dat = dat[, list(data=list(.SD)), by=Region]

# Operations
dat[, model := lapply(data, function(x) lm(Donations ~ Clergy, x))]
dat[, result := lapply(model, tidy)]

# Unnest
dat[, result[[1]], by=Region]
```

## `tidyverse`

The same result can be achieved in a `tidyverse` style:

```{r}
#| message: false
#| warning: false
dat <- "https://vincentarelbundock.github.io/Rdatasets/csv/HistData/Guerry.csv" %>%
       read_csv %>%
       drop_na %>%
       nest_by(Region) %>%
       mutate(model = list(lm(Donations ~ Clergy, data)),
              result = list(tidy(model))) %>%
       unnest(result)
dat
```

## Benchmarking

Admittedly, benchmarks are pretty useless in this context, because unnesting is unlikely to be your major bottleneck unless you are dealing with a ton of subgroups. Still, it is interesting to note that unnesting `data.table` is about 8x faster than `tidyverse` in my naive tests:

```{r}
fit = function(k) lm(y ~ x, k)
dat_dt = data.table(group=sample(1:5000, 1e4, replace=TRUE),
                    y=rnorm(1e4), 
                    x=rnorm(1e4))
dat_dt = dat_dt[, .(data=.(.SD)), by=group][
                , model := lapply(data, fit)][
                , result := lapply(model, tidy)]
dat_tb = tibble(dat_dt)

unnest_dt = function() {
  dat_dt[, result[[1L]], by=group][dat_dt, on='group']
}
unnest_tb = function() {
  dat_dt %>% unnest(result)
}

microbenchmark(unnest_dt(), unnest_tb(), times=5)
```

Note that in the code above, I merged the original dataset back into data.table for a fairer comparison (`tidyr::unnest` returns all the original data).
