---
title = "Calepin vs Quarto"
---

*Calepin* aims for compatibility with [Quarto](https://quarto.org/)'s `.qmd` syntax, but makes different design choices. This page explains the main differences.


# Architecture

Quarto delegates most of its parsing and format conversion to [Pandoc](https://pandoc.org/). This gives Quarto access to Pandoc's many output formats (EPUB, DOCX, ODT, etc.), but it also means that Quarto's rendering speed is bounded by Pandoc's.

*Calepin* does not use Pandoc. Markdown parsing (via comrak), code execution, citation processing, cross-references, and template rendering are all implemented in Rust within a single binary. This makes *Calepin* much faster, but it currently supports fewer output formats: HTML, LaTeX, Typst, and Markdown.


# Performance

Quarto typically renders a document in seconds. *Calepin* renders the same document in milliseconds (excluding code execution time, which is the same in both).

This difference matters in practice. With *Calepin*, the live preview server can re-render and reload the browser instantly every time you save a file. For large websites with hundreds of pages, the difference between seconds-per-page and milliseconds-per-page is the difference between a minutes-long build and one that finishes before you can switch to your browser.

Note that [Quarto 2](https://quarto.org/) is in development and should offer performance in the same class as *Calepin*.


# Configuration: YAML vs TOML

Quarto uses YAML front matter. *Calepin* uses TOML front matter:

```toml
---
title = "My Document"
author = "Jane Doe"
bibliography = "refs.bib"
number-sections = true
---
```

A YAML compatibility layer may be introduced before version 1.0.0.


# Customization philosophy

Quarto provides extensive hard-coded settings in the YAML front matter: hundreds of keys that control layout, styling, and behavior. Changing the look of a Quarto document means finding the right YAML key and hoping it does what you want.

*Calepin* takes a different approach: every piece of output is controlled by a plain-text Jinja template (a "partial"). To change how something looks, you edit the template directly. There are no hidden settings or special YAML keys. The templates are simple HTML, LaTeX, or Typst files with `{{ variable }}` placeholders.

Quarto also uses partials, but they are written in Pandoc's custom template language, which is far less expressive than Jinja2. The Quarto developers are also less committed to exposing users directly to those templates. This is a difference in emphasis, but it matters in the workflow: in *Calepin*, templates are the primary customization mechanism, not a hidden implementation detail.

This design is especially well-suited to working with LLMs. Since templates are just plain text in well-known formats, you can describe what you want and let your LLM agent customize your website, templates, and styles directly:

> Go to the `_calepin/partials` folder. Switch all the colors to a pastel theme. Make the top navigation bar twice as wide. Align the footer text to the right.


# Documentation extraction

*Calepin* includes built-in documentation extraction for R and Python packages via the `calepin man` command. It parses R help files (Rd) and Python docstrings (Google, NumPy, Sphinx, and Markdown styles) and generates `.qmd` reference pages automatically. This makes *Calepin* a natural choice for software documentation websites.


# Syntax differences

## Chunk options

*Calepin* accepts both Quarto's header syntax and `#|` pipe comments. Header key-value pairs are converted internally to pipe-equivalent options:

````
```{r, scatter, echo=FALSE, fig.width=5}
plot(1:10)
````

is equivalent to:

````
```{r, scatter}
#| echo: false
#| fig-width: 5
plot(1:10)
````

When both are present, pipe comments take precedence. Option names with dots (e.g., `fig.width`) are normalized to dashes (`fig-width`). Boolean values like `TRUE`/`FALSE` are normalized to `true`/`false`.

## Labels

In *Calepin*, chunk labels must be in the header. Using `#| label:` in a pipe comment produces an error directing you to move it to the header.

## Extensions directory

Quarto uses `_extensions/` for custom formats and filters. *Calepin* uses `_calepin/partials/` for template overrides and `_calepin/modules/` for custom modules.
