---
title: "What to do with age? Linear, Discrete, Both, or Spline"
author: "Vincent Arel-Bundock"
date: "2024-06-19"
---

In [a blog post](https://statmodeling.stat.columbia.edu/2024/06/19/what/), Andrew @gelman_what_2024 relays this question:

> "The subgroups they examine are 65-74 years old, 75-84, and >85. I’ve seen these types of binning common in medical studies. But why? The actual ages are certainly known, so why not treat age as a continuous variable?"

Gelman's answer is:

> "Discrete binning isn’t perfect, but it’s transparent and can be better than a simple linear model."

The idea is that fitting a model with multiple coefficients associated with bin-specific binary variables allows more flexibility than a model where age is entered purely linearly, with a single coefficient. The downsides of this strategy are that binning "wastes" information by coarsening a fine-grained measure and that the cutpoints of the age bins can sometimes feel arbitrary.

As Gelman notes, however, analysts can get the best of both worlds by including age as *both* a set of bins and as a single continuous variable.

Here, I use the `marginaleffects` package to illustrate four different approaches to modeling age:

1. Linear
2. Discrete
3. Linear and discrete
4. Spline (via a Generalized Additive Model)

First, we load libraries and the Titanic data from [the Rdatasets archive:](https://vincentarelbundock.github.io/Rdatasets/)

```{r}
#| include: false
library(marginaleffects)
library(patchwork)
library(ggplot2)
library(mgcv)
theme_set(theme_bw())
```

```{r} 
library(marginaleffects)
library(patchwork)
library(ggplot2)
library(mgcv)
theme_set(theme_bw())

dat <- "https://vincentarelbundock.github.io/Rdatasets/csv/Stat2Data/Titanic.csv"
dat <- read.csv(dat)
```

Then, we fit 4 different logistic regression models. 

To create bins, we manually enter the cutpoints in the model-fitting formula, using the `I()` syntax. This will allow us to work with `marginaleffects` functions later for plotting and interpretation. For somewhat arbitrary reasons, we use quartiles as cutpoints for our bins:

```{r}
summary(dat$Age)
```

For the model with a spline, we use the `gam()` function from the `mgcv` package:

```{r}
# Linear
m1 <- glm(
    Survived ~ Age,
    family = binomial,
    data = dat)

# Discrete
m2 <- glm(
    Survived ~ I(Age <= 21) + I(Age > 21 & Age <= 30) + I(Age > 30 & Age <= 39),
    family = binomial,
    data = dat)

# Linear & Discrete
m3 <- glm(
    Survived ~ Age + I(Age <= 21) + I(Age > 21 & Age <= 30) + I(Age > 30 & Age <= 39),
    family = binomial,
    data = dat)

# Spline
m4 <- gam(
    Survived ~ s(Age),
    family = binomial,
    data = dat)
```

After fitting the models, we can plot the predicted probability of survival for different ages using the `plot_predictions()` function. As Gelman notes, the

> "fitted model looks kind of goofy: it’s a step function with a slope, so it has a bit of a sawtooth appearance"

```{r}
p1 <- plot_predictions(m1, condition = "Age") + ggtitle("Linear")
p2 <- plot_predictions(m2, condition = "Age") + ggtitle("Discrete")
p3 <- plot_predictions(m3, condition = "Age") + ggtitle("Linear + Discrete")
p4 <- plot_predictions(m4, condition = "Age") + ggtitle("Spline")

p1 + p2 + p3 + p4
```

We can also use the `avg_comparisons()` function to compute the average effect of increasing the `Age` variable by 1 unit on the predicted probability of survival. As you can see, there appear to be meaningful differences in the estimates:

```{r}
avg_comparisons(m1)

avg_comparisons(m2)

avg_comparisons(m3)

avg_comparisons(m4)
```
