---
title: "Consistent figures in Quarto"
author: "Vincent Arel-Bundock"
date: "2024-07-23"
---

I've been struggling to correctly size my `ggplot2` or `base` plots in Quarto documents. The main problem I encounter is that when I change the width or size of a figure, the text in figures also changes in size, which makes the document look very inconsistent.

This post documents [the (awesome) solution](https://r4ds.had.co.nz/graphics-for-communication.html#figure-sizing) I stumbled upon and settled on. There is no explanation, just a series of steps. For discussion, see the links at the bottom of the post.

# Step 1: Defaults

Set reasonable document-wide defaults:

```{r}
#| eval: false
knitr::opts_chunk$set(
    out.width = "70%", # enough room to breath
    fig.width = 6,     # reasonable size
    fig.asp = 0.618,   # golden ratio
    fig.align = "center" # mostly what I want
)
```

# Step 2: Aspect ratio

For each figure, adjust the `fig-asp` code chunk as appropriate. For example:

````{verbatim}
```{r}
#| fig-asp: .7
plot(hp ~ mpg, data = mtcars)
```
````

`fig-asp` is a "free" parameter. You can adjust it as you please without a second thought.

# Step 3: Width

Finally, set the width of your figure. This is a two step process: 

1. Set `out-width`
1. Set `fig-width` based on the value of `out-width`

First, you decide what percentage of the horizontal space you want the figure to take. This is defined in terms of percentage by the `out-width` chunk option. 

Second, use the `out2fig()` calculator to figure out what value of `fig-width` to select:

```{r}
out2fig = function(out.width, out.width.default = 0.7, fig.width.default = 6) {
    fig.width.default * out.width / out.width.default 
}
```

For example, if you use the same defaults as me and want a figure to take up 95% of horizontal space in your document, set the `fig-width` chunk option to:

```{r}
out2fig(0.95)
```

````{verbatim}
```{r}
#| fig-asp: .7
#| out-width: 95%
#| fig-width: 8.142857
plot(hp ~ mpg, data = mtcars)
```
````

# References

* [Figure sizing in R for Data Science](https://r4ds.had.co.nz/graphics-for-communication.html#figure-sizing)
* [Taking control of plot scaling](https://www.tidyverse.org/blog/2020/08/taking-control-of-plot-scaling/)


