---
title: Differences with Quarto
---

*Calepin* aims for compatibility with Quarto's `.qmd` syntax, but there are deliberate differences and simplifications. Most notably, *Calepin* does not use Pandoc. Quarto delegates most of its parsing and format conversion to Pandoc, whereas *Calepin* handles everything internally — markdown parsing (via comrak), code execution, citation processing, cross-references, and template rendering are all built into a single Rust binary. This page documents the differences.


# Chunk headers

In Quarto, the chunk header can contain key-value options:

````
```{r, label, echo=FALSE, fig.width=5}
````

In *Calepin*, the header accepts **only** the language and an optional label:

````
```{r, label}
````

All options must go inside the chunk as `#|` pipe comments:

````
```{r, scatter}
#| echo: false
#| fig-width: 5
````

If you accidentally put key-value pairs in the header, *Calepin* prints an error with the corrected pipe syntax.


# Heading levels

Like Quarto, *Calepin* shifts heading levels down by one when a `title:` is present in the YAML front matter: `#` becomes `<h2>`, `##` becomes `<h3>`, and so on. This is standard practice because the document title occupies `<h1>`. *Calepin* warns if more than one top-level `#` heading is found.

A document without `title:` in the YAML renders headings at their literal level (`#` = `<h1>`).


# Option names

*Calepin* uses dashes in option names (`fig-width`, `fig-cap`), matching Quarto's newer convention. Dots (`fig.width`) are not accepted and will produce a warning suggesting the dashed form.


# Format names

*Calepin* uses canonical format names: `html`, `latex`, `typst`, `markdown`. Aliases like `pdf`, `tex`, `typ`, and `md` are accepted for compatibility with Quarto raw blocks and conditional content (e.g., `::: {.content-visible when-format="pdf"}`).


# Label syntax

In Quarto, a label can be specified either in the header or as a pipe comment:

````
```{r, my-label}
```

```{r}
#| label: my-label
```
````

In *Calepin*, the label must be in the header. Using `#| label:` in a pipe comment produces an error directing you to move it to the header.


# Extensions directory

Quarto uses `_extensions/` for custom formats and filters. *Calepin* uses a project-level directory structure:

```
templates/
  common/         # format-agnostic templates (.jinja)
  html/           # HTML-specific templates
  latex/          # LaTeX-specific templates
  typst/          # Typst-specific templates
snippets/
  common/         # format-agnostic snippets (.jinja)
_calepin/
  plugins/        # plugin directories (filters, custom functions, etc.)
```

The legacy `_calepin/elements/` and `_calepin/templates/` paths are still supported as fallbacks.


# Raw inline content

Inline LaTeX or HTML must be declared as raw content using Quarto's raw attribute syntax, otherwise it will render incorrectly in other output formats. For example, raw LaTeX passed through an HTML render will appear as literal text, and raw HTML in a LaTeX render will break compilation.

Use backtick spans with an explicit format attribute:

```
This has `\textbf{bold}` text.        <!-- BAD: appears literally in HTML -->
This has `\textbf{bold}`{=latex} text. <!-- GOOD: only emitted for LaTeX -->

This has `<mark>highlighted</mark>` text.        <!-- BAD: breaks LaTeX -->
This has `<mark>highlighted</mark>`{=html} text.  <!-- GOOD: only emitted for HTML -->
```

For larger blocks of format-specific content, use a raw fenced div:

````
```{=latex}
\begin{center}
Custom LaTeX here
\end{center}
```

```{=html}
<div class="special">Custom HTML here</div>
```
````

This syntax is identical to Quarto and ensures that format-specific markup is only included in the matching output format.


# Missing features

The following Quarto features are not yet implemented in *Calepin*:

<!-- TODO: fill in as features are tracked -->
