---
title: Differences with Quarto
---

::: {.callout-tip}
Click the `</>` button in the top navigation bar to view the original `.qmd` source code.
:::

*Calepin* aims for compatibility with Quarto's `.qmd` syntax, but there are deliberate differences and simplifications. This page documents them.


# Chunk headers

In Quarto, the chunk header can contain key-value options:

````
```{r, label, echo=FALSE, fig.width=5}
````

In *Calepin*, the header accepts **only** the language and an optional label:

````
```{r, label}
````

All options must go inside the chunk as `#|` pipe comments:

````
```{r, scatter}
#| echo: false
#| fig-width: 5
````

If you accidentally put key-value pairs in the header, *Calepin* prints an error with the corrected pipe syntax.


# Heading levels

Like Quarto, *Calepin* shifts heading levels down by one when a `title:` is present in the YAML front matter: `#` becomes `<h2>`, `##` becomes `<h3>`, and so on. This is standard practice because the document title occupies `<h1>`. *Calepin* warns if more than one top-level `#` heading is found.

A document without `title:` in the YAML renders headings at their literal level (`#` = `<h1>`).


# Option names

*Calepin* uses dashes in option names (`fig-width`, `fig-cap`), matching Quarto's newer convention. Dots (`fig.width`) are not accepted and will produce a warning suggesting the dashed form.


# Format names

*Calepin* uses canonical format names: `html`, `latex`, `typst`, `markdown`. Aliases like `pdf`, `tex`, `typ`, and `md` are accepted for compatibility with Quarto raw blocks and conditional content (e.g., `::: {.content-visible when-format="pdf"}`).


# Label syntax

In Quarto, a label can be specified either in the header or as a pipe comment:

````
```{r, my-label}
```

```{r}
#| label: my-label
```
````

In *Calepin*, the label must be in the header. Using `#| label:` in a pipe comment produces an error directing you to move it to the header.


# Extensions directory

Quarto uses `_extensions/` for custom formats and filters. *Calepin* uses `_calepin/` with a different structure:

```
_calepin/
  elements/       # element templates
  filters/        # executable filters
  templates/      # page templates
  plugins/        # WASM plugins
  shortcodes/     # executable shortcodes
```


# Missing features

The following Quarto features are not yet implemented in *Calepin*:

<!-- TODO: fill in as features are tracked -->
