---
title: "Frontdoor adjustment for causal inference: A primer with examples in `R`"
author: "Vincent Arel-Bundock"
date: "2021-01-24"
---

This notebook has 3 objectives:

1. Offer some intuition about the "frontdoor" criterion and adjustment formula for causal inference.
2. Show how to derive the frontdoor both algebraically and algorithmically using the `dosearch` package for `R`.
3. Show how to apply the frontdoor adjustment formula in simulated data.

I assume that you know what a DAG and a backdoor path are.^[Les lecteurs francophones peuvent se référer au chapitre 6 de mon livre "Analyse Causale et Méthodes Quantitatives", disponible gratuitement en PDF: https://www.pum.umontreal.ca/catalogue/analyse_causale_et_methodes_quantitatives] The data I will consider follows this data generating process, with a cause $X$, a mediator $Z$, an outcome $Y$, and an unobserved confounder $U$:

```{r, fig.width=5, fig.asp=.7, out.width="80%", message=FALSE, echo=FALSE, fig.align = "center"}
library(ggdag)
#theme_set(theme_dag_blank())
co = read.table(header = TRUE, text = "
  x y name
  0 0 X
  1 0 Z
  2 0 Y
  1 1 U")
d = dagify(
  X ~ U,
  Z ~ X,
  Y ~ Z,
  Y ~ U,
  coords = co)
ggdag(d)
```

@Pea2009 uses the $do()$ operator to represent variables on which we intervene or that we manipulate.^[The manipulation could be hypothetical or counterfactual.] For instance, the expression $P(Y|do(X=x))$ represents the distribution of the $Y$ when we manipulate the treatment variable $X$ to give it a specific value $x$.

Our goal is to estimate $P(Y|do(X))$. Unfortunately, this relationship between $X$ and $Y$ is confounded by the unobserved variable $U$, via this backdoor path: $X \leftarrow U \rightarrow Y$. Therefore, we cannot estimate the causal quantity of interest directly.

Instead, we will estimate the effect of $X$ on $Y$ *indirectly* via frontdoor adjustment. The key intuition behind this approach is this:

> In a causal chain with three nodes $X\rightarrow Z\rightarrow Y$, we can estimate the effect of $X$ on $Y$ indirectly by combining two distinct quantities: (1) an estimate of the effect of $X$ on $Z$, and (2) an estimate of the effect of $Z$ on $Y$. 

Roughly speaking, frontdoor adjustment thus proceeds in 3 steps:

1. Estimate $P(Z|do(X))$
1. Estimate $P(Y|do(Z),X)$
1. Combine the two

To illustrate these steps, I will use a simulated dataset that conforms to the DAG above, and where the true effect of $X$ on $Y$ is equal to 0.25:

```{r, message=FALSE}
library(data.table)
set.seed(731460)

N = 1e5
U = rbinom(N, 1, prob = .2)
X = rbinom(N, 1, prob = .1 + U * .6)
Z = rbinom(N, 1, prob = .3 + X * .5)
Y = rbinom(N, 1, prob = .1 + U * .3 + Z * .5)
dat = data.table(X, Z, Y)
```

In the simplest case, we can estimate the effect of $X$ on $Y$ by multiplying two linear regression coefficients. 

Frist we estimate the effect of $Z$ on $X$. Since there is no open backdoor, we do not need to control for other variables:

```{r}
step1 = lm(Z ~ X, dat)
```

Then, we estimate the effect of $Z$ on $Y$, controlling for $X$ to close the backdoor:

```{r}
step2 = lm(Y ~ Z + X, dat)
```

Finally, we combine the two estimates by multiplication:

```{r}
coef(step1)["X"] * coef(step2)["Z"] 
```

Why did this produce the correct result (0.25)? To answer this question, we can use do-calculus to derive a frontdoor adjustment formula, or we can use an `R` package called `dosearch` to derive the formula automatically. The algorithmic approach is very useful in more complicated cases, but the algebraic approach helps us understand the assumptions that underlie the method.

# Algebraic frontdoor

I borrow notation from [a nice Cross Validated answer](https://stats.stackexchange.com/questions/312992/causal-effect-by-back-door-and-front-door-adjustments), and make the following assumptions: 

1. *Full mediation:* there is no directed path from $X$ to $Y$, except through $Z$.
2. *Unconfoundedness 1:* There is no open backdoor from $X$ to $Z$.
3. *Unconfoundedness 2:* All backdoors from Z to Y are blocked by X.

The estimation proceeds in three steps.

*Step 1:* Under assumption 2, the relationship between $X$ and $Z$ *is not* confounded (see DAG at the top). As a result, we have:

$$
P(Z|do(X)) = P(Z|X)
$$ 

*Step 2:* In contrast, the relationship between $Z$ and $Y$ is confounded. Thankfully, adjusting for $X$ suffices to estimate the effect of $Z$ on $Y$, because it blocks the backdoor path.  As a result, we can use the backdoor adjustment formula^[See @Pea2009 for a detailed treatment of backdoor adjustment, and @Pea2016 for an accessible primer.] to get:

$$
P(Y|do(Z)) = \sum_{X}P(Y|X, Z) P(X)
$$

*Step 3:* Back out the effect of $X$ on $Y$ by combining what we obtained above:

$$
\begin{aligned}
P(Y|do(X))  &= \sum_{Z} P(Y|Z, do(X))P(Z|do(X)) && \mbox{(1)}\\
            &= \sum_{Z} P(Y|do(Z))P(Z|do(X))    && \mbox{(2)}\\
            &= \sum_{Z} P(Y|do(Z))P(Z|X)        && \mbox{(3)}\\
            &= \sum_{Z} \sum_{X}P(Y|X, Z) P(X)P(Z|X) && \mbox{(4)}\\
            &= \sum_{Z}P(Z|X) \sum_{X}P(Y|X, Z) P(X) && \mbox{(5)}
\end{aligned}
$$

Equation (1) conditions on Z and sums over its values. Equation (2) is allowed because the effect of $X$ is entirely mediated by $Z$, and because $X$ blocks the backdoor path from $Z$ to $Y$. Intuitively, if we manipulate $Z$, it no longer matters what happened to $X$. Equation (3) is allowed because the relationship between $X$ and $Z$ is unconfounded. Equation (4) applies the backdoor adjustment formula to estimate the effect of $Z$ on $Y$ by conditioning on $X$.

Equation (5) is the frontdoor adjustment formula. The left part is the effect of $X$ on $Z$. The right part is the effect of $Z$ on $Y$.

# Algorithmic frontdoor: `R` & `dosearch`

The `dosearch` package for `R` includes an algorithm that can automatically apply the rules of do-calculus to convert DAGs to adjustment formulas. Instead of manipulating equations ourselves like we did above, we can simply call the `dosearch` function to obtain the frontdoor adjustment formula automatically:

```{r, results="asis"}
library('dosearch')

data1 = "P(X, Y, Z)"

query1 = "P(Y | do(X))"

graph1 = "
U -> X
U -> Y
X -> Z
Z -> Y
"

# compute
frontdoor = dosearch(data1, query1, graph1)

# convert to Rmarkdown equation
cat(paste("$$", frontdoor$formula, "$$"))
```

Which is equivalent to the formula we obtained above.

# Example: Frontdoor simulation

With the frontdoor adjustment formula in hand, we can finally estimate the causal effect of $X$ in our simulated data. We will work with this version fo the formula:

$$
P(Y|do(X)) = \sum_{Z}P(Z|X) \sum_{X}P(Y|X, Z) P(X) 
$$

```{r}

dat[, `P(X)`     := fifelse(X == 1, mean(X), 1 - mean(X)) ][
    , `P(Z|X)`   := mean(Z), by = X                       ][
    , `P(Y|Z,X)` := mean(Y), by = .(Z, X)                 ][
    , `P(Z|X)`   := mean(Z), by = X                       ][
    , Y := NULL                                           ]
dat = unique(dat)
dat[, `P(Y|do(Z))` := sum(`P(Y|Z,X)` * `P(X)`), by = Z]

`P(Y|do(X=0))` = with(dat[X == 0], 
  `P(Z|X)`           [Z == 1] * 
  `P(Y|do(Z))`       [Z == 1] +
  (1 - `P(Z|X)`)     [Z == 0] * 
  `P(Y|do(Z))`       [Z == 0]
)

`P(Y|do(X=1))` = with(dat[X == 1], {
  `P(Z|X)`           [Z == 1] * 
  `P(Y|do(Z))`       [Z == 1] +
  (1 - `P(Z|X)`)     [Z == 0] * 
  `P(Y|do(Z))`       [Z == 0]
})

`P(Y|do(X=1))` - `P(Y|do(X=0))`
```

As shown above, we can get essentially the same result using regression and multiplication:

```{r}
coef(lm(Y ~ Z + X))["Z"] * coef(lm(Z ~ X))["X"]
```

Or by estimating an impossible model (remember that $U$ is unobservable):

```{r}
coef(lm(Y ~ X + U))["X"]
```


